Allee Effect and Cultural Evolution

### 2015-06-05 manuscript/draft.tex
Minor Updates

%\documentclass[review,authoryear]{elsarticle}
 \documentclass[3p,authoryear,twocolumn]{elsarticle} %Just to give a feel of how it would look like

@@

%Check if $R_{max}=\frac{r(A-K)^2}{4AK}$ this actually true, or is rather $R_{max}=\frac{r(K-A)2}{4AK}$

@@

We investigate the equilibrium conditions for different initial population sizes of  $a$ and $b$ and different settings of $z$, $A$, and $K$, assuming that  $r_a=r_b$, and for convenience holding the condition $K_b \geq K_a$. Given that equations \ref{eq1}-\ref{eq4} have no analytical solution, we identified equilibrium conditions via simulation (see appendix 1 for details; source code can be found on the following repository %put link here
as well as in the ESM).  
%if the points we need to reiterate are short, perhaps a footnote will suffice instead of an appendix? 
% 100,000 iterations 
% basin plots resolution

@@

When $z=0$, equations \ref{eq1}-\ref{eq4} \emph{de facto} portray a standard cubic growth model with Allee effect, where the success (or failure) of a given focal trait is exclusively given by its own density. For a generic cultural trait $x$, the model predicts three equilibrium points at $x=0$ (extinction), $x=A_x$, and $x=K_x$. The equilibrium $x=A_x$ is unstable in all cases, whereas the stability of the node at $x=K_x$ is a function of the per capita growth rate $r_x$. Increasing the latter decreases the stability of fixed points~\citep{scheuring_1999}, leading to the transition from point attractors (with equilibrium at $x=K_x$), to limit cycles and chaotic dynamics with islands of stability (see fig. \ref{fig:bifurcationDiagram}).

\begin{figure}[h!]
  \centering
      \includegraphics[width=0.5\textwidth]{./figures/figure1.jpg}
  \caption{Bifurcation diagram of a single population model with $A_x=0.2$ and $K_x=0.8$. For $0<r_x<0.6649747$, any initial value of $x$ greater than $A_x$ converges to $K_x$.}
    \label{fig:bifurcationDiagram}
\end{figure}

Here we consider only scenarios where the node $x=K_x$ is stable. In this case the initial value of $x(t=0)$ will determine the ultimate equilibria of the system: extinction, if $x(t=0)<A_x$; unstable equilibrium, if $x(t=0)=A_x$; and stable equilibrium, if $x(t=0)>A_x$. Thus in a 2-trait model, the coexistence of both variants is guaranteed only when the initial density of both traits are above their respective critical densities, otherwise we should expect that either (i.e. $a(t=0) \geq A_a$ and $b(t=0)<A_b$ or $a(t=0)<A_a$ and $b(t=0) \geq A_b$) or both (i.e. when $a(t=0)<A_a$ and $b(t=0)<A_b$) variants are ultimately extinct (see fig. \ref{fig:NoTransmissionBasin}).

\begin{figure}[h!]
  \centering
      \includegraphics[width=0.5\textwidth]{./figures/figure2.pdf}
  \caption{Basins of attraction for the four equilibria (mutual extinction: grey; extinction of $a$: blue; extinction of $b$: red; stable coexistence: green). Solid lines represent trajectories of the system for different initial conditions, represented by dots. Model parameters: $A_a=0.2$; $K_a=0.7$; $A_b=0.5$; $K_b=1.0$; $r_a=r_b=0.05$}.
    \label{fig:NoTransmissionBasin}
\end{figure}

>>>>

\documentclass[review,authoryear]{elsarticle}
% \documentclass[3p,authoryear,twocolumn]{elsarticle} %Just to give a feel of how it would look like

@@

@@

We investigate the equilibrium conditions for different initial population sizes of $a$ and $b$ and different settings of $z$, $A$, and $K$, ensuring that in all cases $K_b \geq K_a$. We assumed that two strategies differed only in their critical densities and carrying capacities but not on their intrinsic payoffs. To achieve this we set $r_b=\rho r_a$, where $\rho$ is a constant\footnote{Given two variants with the same baseline growth rate (i.e. $r_a=r_b$), equation \eqref{eq2} result into a lower $R$ for the variant with higher $A$ and $K$. In order to avoid this we increased the baseline of the strategy with lower $R$ by a factor $\rho$, given by the following equation:
\begin{align*}
\rho=\frac{A_bK_b(K_a-A_a)^2}{A_aK_a(A_b-K_b)^2}
\end{align*}
  } 
that ensures that the two variants have always the same $R_{max}$. Given that equations \ref{eq1}-\ref{eq4} have no analytical solution, we identified equilibrium conditions via simulation (source codes of the simulation can be found on the following repository \textcolor{red}{PUT LINK HERE})

@@

When $z=0$, equations \ref{eq1}-\ref{eq4} becomes a standard cubic growth model with Allee effect, where the success (or failure) of a given focal trait is exclusively given by its own density. For a generic cultural trait $x$, the model predicts three equilibrium points at $x=0$ (extinction), $x=A_x$, and $x=K_x$. The equilibrium $x=A_x$ is unstable for all values of $r_x$, whereas the stability of the node at $x=K_x$ is a function of the per capita growth rate $r_x$. Increasing the latter decreases the stability of fixed points~\citep{scheuring_1999}, leading to the transition from point attractors (with equilibrium at $x=K_x$), to limit cycles and chaotic dynamics with islands of stability (see fig. \ref{fig:bifurcationDiagram}).

\begin{figure}[h!]
  \centering
      \includegraphics[width=1\textwidth]{./figures/figure1.jpg}
  \caption{Bifurcation diagram of a single population model with $A_x=0.2$ and $K_x=0.8$. For $0<r_x<0.6649747$, any initial value of $x$ greater than $A_x$ converges to $K_x$.}
    \label{fig:bifurcationDiagram}
\end{figure}

For the rest of the paper we will consider only scenarios where the node $x=K_x$ is stable. In this case the initial value of $x(t=0)$ will determine the ultimate equilibria of the system: extinction, if $x(t=0)<A_x$; unstable equilibrium, if $x(t=0)=A_x$; and stable equilibrium, if $x(t=0)>A_x$. Thus in a 2-trait model, the coexistence of both variants is guaranteed only when the initial density of both traits are above their respective critical densities, otherwise we should expect that either (i.e. $a(t=0) \geq A_a$ and $b(t=0)<A_b$ or $a(t=0)<A_a$ and $b(t=0) \geq A_b$) or both (i.e. when $a(t=0)<A_a$ and $b(t=0)<A_b$) variants are ultimately extinct (see fig. \ref{fig:NoTransmissionBasin}).

\begin{figure}[h!]
  \centering
      \includegraphics[width=1\textwidth]{./figures/figure2.pdf}
  \caption{Basins of attraction for the four equilibria (mutual extinction: grey; extinction of $a$: blue; extinction of $b$: red; stable coexistence: green). Solid lines represent trajectories of the system for different initial conditions, represented by dots. Model parameters: $A_a=0.2$; $K_a=0.7$; $A_b=0.5$; $K_b=1.0$; $r_a=r_b=0.05$}.
    \label{fig:NoTransmissionBasin}
\end{figure}
